% --- LaTeX Essay Template - S. Venkatraman ---

% --- Set document class and font size ---

\documentclass[letterpaper, 11pt]{article}

% --- Package imports ---

% lipsum is just for generating placeholder text and can be removed
\usepackage{hyperref, lipsum} 

% --- Fonts (Set to Palatino or Times New Roman if desired) ---

% \usepackage{newpxtext, newpxmath}
% \usepackage{times}

% --- Page layout settiings ---

% Set page margins
\usepackage[left=1.2in, right=1.2in, top=.9in, bottom=.9in]{geometry}

% Anchor footnotes to the bottom of the page
\usepackage[bottom]{footmisc}
\usepackage{setspace}
\setstretch{1.15}
% Set line spacing
\renewcommand{\baselinestretch}{1.2}

% Use this for a one-off '*' footnote (e.g., a URL in a heading)
\newcommand{\starfootnote}[1]{%
\begingroup
\renewcommand{\thefootnote}{\fnsymbol{footnote}}%
\footnote{#1}%
\endgroup
\addtocounter{footnote}{-1}%
}

% Set spacing between paragraphs
\setlength{\parskip}{2mm}

% --- Page formatting settings ---



% --- Essay title and author ---
\title{\vspace{-1.5cm} Summary of \textit{Science and Statistics}\footnote{\url{https://www.jstor.org/stable/2286841}}}
\author{Shizhe Zhang (3041882158)\\\href{mailto:shizhe\_zhang@berkeley.edu}{shizhe\_zhang@berkeley.edu}}

\date{\today}

% --- Main text ---
% --- Document starts here ---

\begin{document}
% Set link colors for labeled items and URLs (blue) 
\hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue}
\maketitle

Science is an iteration between theory and data.
And statistics is tools that help us learn from those data, but it is not about finding the absolute truth.
All models are simplifications, only to help us understand the world better.
Fisher is the pioneer of statistics, and we should learn from his attitude towards data in our lives.
We should always be skeptical about models and question ourselves about the underlying mechanism behind the data.

Statistics is not mathematics; it is a science that requires subject-matter knowledge, evolving with new methods.
Statistics is not a set of formal procedures, either; it requires judgment and context.

The world is growing more complex, and data are becoming more abundant.
Statistician need to adapt to the changing world and meet new challenges.


% --- Document ends here ---
\end{document}